Cost of Living in Toronto

### 2024-09-22 paper/paper.qmd
added all the necessary graphs in paper.qmd

# Data
```{r}
#| label: low-income-families
#| fig-cap: "Table 1: Sample of the Income/Expenses of Low Income Households in Toronto"
#| echo: false

# PREVIEW OF THE DATA
cleaned_data[, 1:5] %>% head() %>% knitr::kable()
```
## Data Source

## Overview of the Dataset

## Average Income of the 11 Households

## Average Expenses of the 11 Households

## Difference between Income and Expense

# Results

# Discussion

>>>>

# Data

This report uses the "About Cost of living in Toronto for low-income households" (CITATION HERE) dataset, which is published publicly by the City of Toronto on the Open Data Portal. This data contains information on income such as employment income, GST/HST credits, pension benefits and expenses such as rent, food, transportations for 11 different scenarios of a low-income household. The 11 different scenarios of low-income household were manually selected by the City of Toronto to reflect different household characteristics that can increase the reliance on social assistance, such as type of income, single-parent household, etc. A detailed explanation of this 11 household scenarios can be found in the Appendix.

This dataset was downloaded and analyzed using the R Programming Language (R Core Team 2022), opendatatoronto (CITATION HERE), knitr (CITATION HERE), tidyverse (CITATION HERE), here (CITATION), dplyr (CITATION) and sf (CITATION HERE).

## Overview of the Dataset
```{r}
#| label: low-income-families
#| fig-cap: "Table 1: Sample of the Income/Expenses of Low Income Households in Toronto"
#| echo: false

# PREVIEW OF THE DATA
cleaned_data[, 1:5] %>% head() %>% knitr::kable()
```

## Average Income of the 11 Households
```{r}
#| label: low-income-families
#| fig-cap: "Table 2: Average Income of Low Income Households in Toronto"
#| echo: false

avg_income <- filter(cleaned_data, Category == "Income")

# take the average
avg_income$"Average Income in $" <- rowMeans(subset(avg_income, select = c("Household_1",
                                                                           "Household_2",
                                                                           "Household_3",
                                                                           "Household_4",
                                                                           "Household_5",
                                                                           "Household_6",
                                                                           "Household_7",
                                                                           "Household_8",
                                                                           "Household_9",
                                                                           "Household_10",
                                                                           "Household_11"
                                                                           )))
avg_income <- avg_income %>% select("Item", "Average Income in $")

# rename the columns to something more presentable
colnames(avg_income) <- c("Type of Income", "Average Income in $")

#round to the nearest integer
avg_income$"Average Income in $" <- round(avg_income$"Average Income in $", 0)

avg_income %>% head() %>% knitr::kable()
```

## Average Expenses of the 11 Households
```{r}
#| label: low-income-families
#| fig-cap: "Table 3: Average Expenses of Low Income Households in Toronto"
#| echo: false

avg_expense <- filter(cleaned_data, Category == "Employment Deduction" |
                                    Category == "Selected Expenses")

# take the average of the expenses
avg_expense$"Average Expense in $" <- rowMeans(subset(avg_expense, select = c("Household_1",
                                                                           "Household_2",
                                                                           "Household_3",
                                                                           "Household_4",
                                                                           "Household_5",
                                                                           "Household_6",
                                                                           "Household_7",
                                                                           "Household_8",
                                                                           "Household_9",
                                                                           "Household_10",
                                                                           "Household_11"
                                                                           )))
avg_expense <- avg_expense %>% select("Item", "Average Expense in $")

# rename the columns to something more presentable
colnames(avg_expense) <- c("Type of Income", "Average Expense in $")

#round to the nearest integer
avg_expense$"Average Expense in $" <- round(avg_expense$"Average Expense in $", 0)

# absolute value (because employment insurance is relative to income)
avg_expense$"Average Expense in $" <- abs(avg_expense$"Average Expense in $") 
avg_expense %>% head() %>% knitr::kable()

```
## Difference between Income and Expense
```{r}
#| label: low-income-families
#| fig-cap: "Table 4: Average Expenses of Low Income Households in Toronto"
#| echo: false

income_expense_merged <- cleaned_data_inverted

income_expense_merged <- income_expense_merged %>% select("Total Income", "Total Expenses", "Remaining Funds")

income_expense_merged  %>% knitr::kable()
```

# Results
```{r}
#| label: low-income-families
#| fig-cap: "Table 5: Average Expenses of Low Income Households in Toronto"
#| echo: false

data <- data.frame(
  Household = c("Household_1", "Household_2", "Household_3", "Household_4", 
                "Household_5", "Household_6", "Household_7", "Household_8", 
                "Household_9", "Household_10", "Household_11"),
  Income = income_expense_merged$"Total Income",
  Expenses = income_expense_merged$"Total Expenses"
)

data_long <- data %>%
  pivot_longer(cols = c(Income, Expenses), 
               names_to = "Type", 
               values_to = "Amount")

ggplot(data_long, aes(x = Household, y = Amount, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Total Income vs Total Expenses for Each Household", 
       x = "Household", 
       y = "Amount") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r}
#| label: low-income-families
#| fig-cap: "Table 6: Average Expenses of Low Income Households in Toronto"
#| echo: false

expense_disparity <- head(avg_expense, 6)
avg_expense
pie(expense_disparity$'Average Expense in $', expense_disparity$'Type of Income')

```
# Discussion
